quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space

### 2024-03-31 paper-quollr.Rmd
Add dataset for application

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE)

@@

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)

set.seed(20230531)
```

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      echo = FALSE)

@@

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)

set.seed(20240110)
```

```{r plot-theme}
theme_set(theme_linedraw() +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "none", 
     legend.title = element_text(size=5), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm')
   )

)
```

@@

```{r, echo=FALSE}
clusters_different_shapes_diff_num_points <- function(sample_size = 400, with_seed = NULL, cluster_size_vec = c(50, 50, 50, 50, 100, 100), num_gussian_clusters = 4, num_non_gaussian_clusters = 2,
                                      cluster_sd_gau = 0.05, cluster_sd_non_gau = 0.1, num_dims = 7, a = 2, b = 4) {

  # To check the seed is not assigned
  if (!is.null(with_seed)) {
    set.seed(with_seed)
  }

  num_clusters <- num_gussian_clusters + num_non_gaussian_clusters

  ## Generate Gaussian clusters

  # Create a vector of possible values (0 and 1)
  values <- c(0, 1)

  # Create an expanded grid with 0's and 1's
  mean_val_grid <- tidyr::expand_grid(!!!setNames(rep(list(values), num_dims),
                                                  paste0("mean_dim", 1:num_dims)))

  # To select combinations for assigned number of clusters

  mean_val_grid_gau <- mean_val_grid |>
    dplyr::slice_sample(n = num_gussian_clusters)

  mean_val_grid_non_gau <- mean_val_grid |>
    dplyr::slice_sample(n = num_non_gaussian_clusters)

  # To generate empty tibble
  column_names <- paste0(rep("x", num_dims), 1:num_dims)
  df <- tibble(!!!setNames(rep(list(NULL), length(column_names)), column_names))

  for (i in 1:num_gussian_clusters) {

    # To filter the mean values for specific cluster
    mean_val_for_cluster <- mean_val_grid_gau |>
      dplyr::filter(dplyr::row_number() == i) |>
      unlist(use.names = FALSE)

    # Initialize an empty list to store the vectors with column
    # values
    dim_val_list <- list()

    for (j in 1:num_dims) {

      dim_val_list[[column_names[j]]] <- rnorm(cluster_size_vec[i], mean = mean_val_for_cluster[j],
                                               sd = cluster_sd_gau)

    }
    # To generate a tibble for a cluster
    df_gau_cluster <- tibble::as_tibble(dim_val_list)

    df <- dplyr::bind_rows(df, df_gau_cluster)

  }

  for (i in 1:num_non_gaussian_clusters) {

    phi <- runif(cluster_size_vec[(num_clusters - i)], max = 2*pi)
    rho <- sqrt(runif(cluster_size_vec[(num_clusters - i)]))

    # To filter the mean values for specific cluster
    presence_of_elipse_cluster <- mean_val_grid_non_gau |>
      dplyr::filter(dplyr::row_number() == i) |>
      unlist(use.names = FALSE)

    # Initialize an empty list to store the vectors with column
    # values
    dim_val_list_n <- list()

    for (j in 1:num_dims) {
      if(presence_of_elipse_cluster[j] == 1){
        dim_val_list_n[[column_names[j]]] <- sqrt(a)*rho*cos(phi) + b
        ## Surface of poolar coordinate
      } else {
        dim_val_list_n[[column_names[j]]] <- rnorm(cluster_size_vec[(num_clusters - i)], mean = 0,
                                                   sd = cluster_sd_non_gau)

      }

    }
    # To generate a tibble for a cluster
    df_non_gau_cluster <- tibble::as_tibble(dim_val_list_n)

    df <- dplyr::bind_rows(df, df_non_gau_cluster)

  }

  df

}
```

```{r}
clust_df <- clusters_different_shapes_diff_num_points(sample_size = 1500, with_seed = NULL, cluster_size_vec = c(250, 150, 150, 150, 350, 450), num_gussian_clusters = 4, num_non_gaussian_clusters = 2, cluster_sd_gau = 0.2, cluster_sd_non_gau = 0.3, num_dims = 7, a = 2, b = 4)

langevitour::langevitour(clust_df)

clust_df <- clust_df |>
  dplyr::mutate(ID = row_number())
```

```{r}
library(Rtsne)

tSNE_fit <- clust_df |>
  dplyr::select(-ID) |>
  dplyr::select(where(is.numeric)) |>
  Rtsne::Rtsne(perplexity = 39, pca = FALSE, pca_center = FALSE, normalize = FALSE, dims = 2)

tSNE_df <- tSNE_fit$Y |> 
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::mutate(ID = clust_df$ID)

names(tSNE_df) <- c("tSNE1", "tSNE2", "ID")

tSNE_df_plot <- tSNE_df |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.5, colour="#e41a1c", size = 0.5) +
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.text = element_text(size = 5),
        axis.title = element_text(size = 7))

tSNE_df_plot
```

<!-- MSE plot to find which is the effective number of bins to use-->
```{r}
tsne_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tSNE_df, 
                                    x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tSNE_df))    
## tSNE
hex_size_vec <- seq(0.02, 2, by = 0.01)
vec <- stats::setNames(rep("", 6), c("num_bins", "aic", "mse", "num_bins_x", "num_bins_y", "hex_size"))  ## Define column names
mse_df_gau <- dplyr::bind_rows(vec)[0, ]
mse_df_gau <- mse_df_gau |>
  dplyr::mutate_if(is.character, as.numeric)
for (i in 1:length(hex_size_vec)) {

  num_bin_list <- calc_bins(data = tsne_gau_scaled, 
            x = "tSNE1", y = "tSNE2", 
            hex_size = hex_size_vec[i], buffer_x = NA, buffer_y = NA)

  num_bins_x <- num_bin_list$num_x
  num_bins_y <- num_bin_list$num_y

  model_object <- fit_highd_model( training_data = clust_df, 
                                   nldr_df_with_id = tsne_gau_scaled, 
                                   x = "tSNE1", y = "tSNE2", 
                                   num_bins_x = num_bins_x, 
                                   num_bins_y = num_bins_y, 
                                   x_start = NA, y_start = NA, 
                                   buffer_x = NA, buffer_y = NA, 
                                   hex_size = hex_size_vec[i],
                                   is_rm_lwd_hex = FALSE, 
                                   benchmark_to_rm_lwd_hex = NA, 
                                   col_start_2d = "tSNE", 
                                   col_start_highd = "x")

  centroid_df_training <- model_object$df_bin_centroids
  avg_df_training <- model_object$df_bin

  pred_emb_list <- predict_emb(test_data = clust_df, 
                                  df_bin_centroids = centroid_df_training, 
                                  df_bin = avg_df_training, type_NLDR = "tSNE")

  pred_df_training <- as.data.frame(do.call(cbind, pred_emb_list))

  eval_list <- gen_summary(test_data = clust_df, 
                                  prediction_df = pred_df_training, 
                                  df_bin = avg_df_training, col_start = "x")

  mse_df_gau <- mse_df_gau |>
    tibble::add_row(num_bins = num_bins_x * num_bins_y,
                    aic = eval_list$aic,
                    mse = eval_list$mse,
                    num_bins_x = num_bins_x,
                    num_bins_y = num_bins_y,
                    hex_size = hex_size_vec[i])

}
## If same total number of bins occurred only select ones with minimum error
### Obtain duplicate bins
dupli_bins <- mse_df_gau |> 
  dplyr::count(num_bins) |> 
  dplyr::filter(n > 1) |> 
  dplyr::pull(num_bins)
### Group split by duplicated bins
duplicate_df_list <- mse_df_gau |>
  dplyr::filter(num_bins %in% dupli_bins) |>
  dplyr::arrange(num_bins) |>
  dplyr::group_split(num_bins)
### Obtain one row from duplicates which have lowest error and hexsize
duplicate_df <- data.frame(matrix(nrow = 0, ncol = 0))
for (i in 1:length(duplicate_df_list)) {

  dd <- duplicate_df_list[[i]] |>
    dplyr::filter(mse == min(duplicate_df_list[[i]]$mse)) |>
    dplyr::filter(hex_size == min(duplicate_df_list[[i]]$hex_size))

  duplicate_df <- dplyr::bind_rows(duplicate_df, dd)

}
### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df_gau |>
  dplyr::filter(!(num_bins %in% dupli_bins))
### Combine duplicated and not duplicated(corrected) bins dfs
mse_df_clust <- dplyr::bind_rows(not_dupli_df, duplicate_df) 
```

```{r}
mse_clust_plot <- ggplot(mse_df_clust, aes(x = num_bins,
                                       y = log(mse)
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 588, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  ylab("MSE") +
  xlab("total number of bins")

mse_clust_plot
```

<!-- model with bin centroids-->
```{r}

## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 21
num_bins_y_tsne_gau <- 28
hex_size_tsne_gau <- 0.03
## non-empty:198
hb_obj_tsne_gau <- hex_binning(data = tsne_gau_scaled, x = "tSNE1", 
                      y = "tSNE2", num_bins_x = num_bins_x_tsne_gau, 
                      num_bins_y = num_bins_y_tsne_gau, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_tsne_gau, col_start = "tSNE")
## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$centroids))
## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$hex_poly))
## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
# ggplot(data = hex_grid, aes(x = x, y = y)) +
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
#   coord_fixed()
hex_grid_with_counts <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hb_id"))
# ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
#   geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
#   geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID)) +
#   scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
#   coord_fixed()
tsne_data_with_hb_id_gau <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$data_hb_id))

model_object_tsne_gau <- fit_highd_model( training_data = clust_df, 
                                 nldr_df_with_id = tsne_gau_scaled, 
                                 x = "tSNE1", y = "tSNE2", 
                                 num_bins_x = num_bins_x_tsne_gau, 
                                 num_bins_y = num_bins_y_tsne_gau, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_tsne_gau,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "tSNE", 
                                 col_start_highd = "x")
df_bin_centroids_tsne_gau <- model_object_tsne_gau$df_bin_centroids
df_bin_tsne_gau <- model_object_tsne_gau$df_bin
## Triangulate bin centroids
tr1_object_tsne_gau <- tri_bin_centroids(df_bin_centroids_tsne_gau, x = "c_x", y = "c_y")
tr_from_to_df_tsne_gau <- gen_edges(tri_object = tr1_object_tsne_gau)
## Compute 2D distances
distance_tsne_gau <- cal_2d_dist(tr_coord_df = tr_from_to_df_tsne_gau, 
                             start_x = "x_from", start_y = "y_from", 
                             end_x = "x_to", end_y = "y_to", 
                             select_vars = c("from", "to", "distance"))
# distance_plot <- plot_dist(distance_tsne_gau) +
#   ylab(expression(d^{(2)})) +
#   theme(axis.text = element_text(size = 5),
#         axis.title = element_text(size = 12))
# 
# distance_plot
## To find the benchmark value
benchmark_tsne_gau <- find_lg_benchmark(distance_edges = distance_tsne_gau, distance_col = "distance")
# ggplot() +
# geom_trimesh(data = df_bin_centroids_tsne_gau, mapping = aes(x = c_x, y = c_y))
trimesh_removed_tsne_gau <- vis_rmlg_mesh(distance_edges = distance_tsne_gau, benchmark_value =
benchmark_tsne_gau, tr_coord_df = tr_from_to_df_tsne_gau, distance_col = "distance")
trimesh_removed_tsne_gau <- trimesh_removed_tsne_gau +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

trimesh_removed_tsne_gau

df_all_tsne_gau <- dplyr::bind_cols(clust_df |> dplyr::select(-ID), tsne_data_with_hb_id_gau)

### Define type column
df <- df_all_tsne_gau |>
  dplyr::select(tidyselect::starts_with("x")) |>
  dplyr::mutate(type = "data") ## original dataset

df_b <- df_bin_tsne_gau |>
  dplyr::filter(hb_id %in% df_bin_centroids_tsne_gau$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean
## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_tsne_gau$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)
df_exe <- dplyr::bind_rows(df_b, df)
## Set the maximum difference as the criteria
distance_df_small_edges <- distance_tsne_gau |>
  dplyr::filter(distance < benchmark_tsne_gau)
## Since erase brushing is considerd.
langevitour::langevitour(df_exe[1:(length(df_exe)-1)],
                         lineFrom = distance_df_small_edges$from,
                         lineTo = distance_df_small_edges$to,
                         group = df_exe$type, pointSize = append(rep(0, NROW(df_b)), rep(0.7, NROW(df))),
                         levelColors = c("#6a3d9a", "#33a02c"))

<!-- model with bin means-->
```{r}
## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 21
num_bins_y_tsne_gau <- 28
hex_size_tsne_gau <- 0.03
## non-empty:198
hb_obj_tsne_gau <- hex_binning(data = tsne_gau_scaled, x = "tSNE1", 
                      y = "tSNE2", num_bins_x = num_bins_x_tsne_gau, 
                      num_bins_y = num_bins_y_tsne_gau, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_tsne_gau, col_start
= "tSNE")
## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$centroids))
## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$hex_poly))
## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
# ggplot(data = hex_grid, aes(x = x, y = y)) +
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
#   coord_fixed()
hex_grid_with_counts <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" =
"hb_id"))
# ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
#   geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
#   geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID)) +
#   scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
#   coord_fixed()
tsne_data_with_hb_id_gau <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$data_hb_id))

model_object_tsne_gau <- fit_highd_model( training_data = clust_df, 
                                 nldr_df_with_id = tsne_gau_scaled, 
                                 x = "tSNE1", y = "tSNE2", 
                                 num_bins_x = num_bins_x_tsne_gau, 
                                 num_bins_y = num_bins_y_tsne_gau, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_tsne_gau,
                                 is_bin_centroid = FALSE,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "tSNE", 
                                 col_start_highd = "x")
df_bin_centroids_tsne_gau <- model_object_tsne_gau$df_bin_centroids
df_bin_tsne_gau <- model_object_tsne_gau$df_bin
## Triangulate bin centroids
tr1_object_tsne_gau <- tri_bin_centroids(df_bin_centroids_tsne_gau, x = "c_x", y = "c_y")
tr_from_to_df_tsne_gau <- gen_edges(tri_object = tr1_object_tsne_gau)
## Compute 2D distances
distance_tsne_gau <- cal_2d_dist(tr_coord_df = tr_from_to_df_tsne_gau, 
                             start_x = "x_from", start_y = "y_from", 
                             end_x = "x_to", end_y = "y_to", 
                             select_vars = c("from", "to", "distance"))
# distance_plot <- plot_dist(distance_tsne_gau) +
  # ylab(expression(d^{(2)})) +
  # theme(axis.text = element_text(size = 5),
        # axis.title = element_text(size = 12))

# distance_plot
## To find the benchmark value
benchmark_tsne_gau <- find_lg_benchmark(distance_edges = distance_tsne_gau, distance_col =
"distance")

benchmark_tsne_gau <- 0.17

# ggplot() +
# geom_trimesh(data = df_bin_centroids_tsne_gau, mapping = aes(x = c_x, y = c_y))
trimesh_removed_tsne_gau <- vis_rmlg_mesh(distance_edges = distance_tsne_gau, 
                                          benchmark_value = benchmark_tsne_gau, 
                                          tr_coord_df = tr_from_to_df_tsne_gau, distance_col = "distance")
trimesh_removed_tsne_gau <- trimesh_removed_tsne_gau +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size =
3)
trimesh_removed_tsne_gau
df_all_tsne_gau <- dplyr::bind_cols(clust_df |> dplyr::select(-ID),
tsne_data_with_hb_id_gau)
### Define type column
df <- df_all_tsne_gau |>
  dplyr::select(tidyselect::starts_with("x")) |>
  dplyr::mutate(type = "data") ## original dataset
df_b <- df_bin_tsne_gau |>
  dplyr::filter(hb_id %in% df_bin_centroids_tsne_gau$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean
## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_tsne_gau$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)
df_exe <- dplyr::bind_rows(df_b, df)
## Set the maximum difference as the criteria
distance_df_small_edges <- distance_tsne_gau |>
  dplyr::filter(distance < benchmark_tsne_gau)
## Since erase brushing is considerd.
langevitour::langevitour(df_exe[1:(length(df_exe)-1)],
                         lineFrom = distance_df_small_edges$from,
                         lineTo = distance_df_small_edges$to,
                         group = df_exe$type, pointSize = append(rep(0, NROW(df_b)),
rep(0.7, NROW(df))),
                         levelColors = c("#6a3d9a", "#33a02c"))
```

### 2024-03-31 paper-quollr.tex
Add dataset for application

\begin{verbatim}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)

set.seed(20230531)
\end{verbatim}

@@

>>>>

@@

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-2-1.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-3-1.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-5-1.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-6-1.pdf} \includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-6-2.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-7-1.pdf} \includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-7-2.pdf}
